European or American? A re-examination of the transatlantic influence over the IMF

### 2025-09-18 manuscript/manuscript.qmd

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
date-format: |
    | DD MMMM YYYY
authors:
  - name: Dianyi Yang
    email: dianyi.yang@politics.ox.ac.uk
    orcid: 0009-0004-4652-3429
    corresponding: true
    affiliations:
      - ref: ox    
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
    affiliations:
      - ref: bw
    equal-contributor: true
  - name: Zhenyi Chen
    orcid: 0009-0008-2910-5187
    affiliations:
      - ref: rmu
    equal-contributor: true
affiliations:
  - id: ox
    name: Department of Politics and International Relations, University of Oxford
  - id: bw
    name: School of International Relations and Diplomacy, Beijing Foreign Studies University
  - id: rmu
    name: School of International Studies, Renmin University of China
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
          \AtAppendix{\counterwithin{table}{section}}
abstract: "This paper investigates whether the United States or Europe holds greater influence over the International Monetary Fund (IMF)—a longstanding yet underexamined question in the study of global economic governance. While the Fund is widely regarded as dominated by “Western” powers, most existing research either focuses exclusively on American influence or treats the West as a unified bloc, masking important variation between its two key actors. Drawing on panel data from 1980 to 2010, we construct separate measures of US and European influence using principal component analysis of commonly used alignment indicators—UN voting patterns, bilateral trade, and banking exposure—and estimate a series of Tobit and Probit models to assess their impact on four key IMF lending outcomes: loan-to-GDP ratios, participation rates, loan approvals, and conditionality. We find that European influence is strongly associated with higher loan volumes, more frequent participation, and greater likelihood of loan approval, while US influence is not. However, American ties are significantly correlated with fewer conditions attached to loans. Additional analysis shows that European influence is not driven by regional or colonial favouritism: neither European nor African borrowers benefit more from European ties—and may, in fact, fare worse—than other countries with comparable alignment. These findings complicate existing assumptions about IMF decision-making, highlighting the need to distinguish between types of Western influence. They also suggest that Europe wields broader institutional influence over access to IMF resources, while US power manifests more selectively through conditionality. In doing so, the paper contributes to debates on hegemony, institutional legitimacy, and intra-Western dynamics in global economic governance."
keywords: International Monetary Fund, hegemony, global economic governance, US influence, European influence, conditionality
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

@@

In principle, quantitative approaches are well suited for directly comparing the influence of different great powers within international organisations. In the IMF context, however, most such studies focus on American influence, with little systematic attention paid to European power. These studies typically proxy the strategic interests of major powers using indicators such as UN voting alignment [@thacker1999; @dreher2009], bilateral trade intensity [@barro2005], and financial exposure [@oatley2004]. Yet, much of this literature has remained US-centric, often using only American alignment as the independent variable of interest [@thacker1999; @oatley2004; @dreher2009]. Other studies pool European indicators with US ones under the broader category of “Western” influence, thus masking potential differences between the two [@lipscy2018; @barro2005]. These ties have been shown to correlate with more favourable IMF lending outcomes—larger loans, higher approval rates, and fewer or less stringent conditions [@lipscy2018; @barro2005; @dreher2009]. Yet, despite the availability of these tools, comparative assessments between the US and Europe remain largely absent from this literature.

@@

-   UN General Assembly voting similarity[^4]
-   Bilateral trade intensity[^5]
-   Banking sector exposure[^6]

[^4]: Originallly used in @thacker1999.

[^5]: From @barro2005.

[^6]: See @broz2006; @lipscy2017; @oatley2004.

```{r percentage_variance}
load('save/PCA.RData')
US_PC1_percentage <- US$sdev[1]^2 / sum(US$sdev^2) * 100
EU_PC1_percentage <- EU$sdev[1]^2 / sum(EU$sdev^2) * 100
```

However, as noted by @lipscy2018 and @belsley2004, including multiple proxies for the same latent construct in a single regression model leads to multicollinearity, which reduces statistical power and inflates standard errors, as the variables are highly correlated in our data[^7]. To address this issue, we apply principal component analysis (PCA) to reduce the dimensionality of the data [@dunteman2008]. Specifically, we extract the first principal component — a linear combination of the original variables that captures the greatest share of variance — to serve as a single measure of influence[^8].

[^7]: See Appendix [-@sec-corr].

[^8]: The first principal components capture the majority of the variance (`r format(EU_PC1_percentage, digits = 1)`% for Europe and `r format(US_PC1_percentage, digits = 1)`% for the United States). See Appendix [-@sec-scree] for scree plots and additional details.

@@

@@

[^10]: Quadratic terms are included for GDP and per capita GDP to capture potential non-linear effects. See [Model Specification](#sec-model) for details.

## Summary Statistics

@@

@@

This paper has examined a fundamental yet underexplored question in the study of global economic governance: who wields greater influence over the International Monetary Fund—the United States or Europe? By disaggregating "Western" influence into its constituent components and employing principal component analysis to construct separate measures of US and European alignment, we have uncovered important asymmetries in how transatlantic powers exercise their influence within the Fund.

Our empirical analysis yields several key findings that challenge conventional wisdom about IMF decision-making. First, European influence demonstrates a broader and more consistent association with favourable lending outcomes than previously recognised. Countries with stronger ties to major European powers，namely, the United Kingdom, France, and Germany, receive significantly larger loans relative to GDP, participate more frequently in IMF programmes, and enjoy higher approval rates for new lending arrangements. These effects are statistically significant and substantively meaningful, suggesting that European actors wield considerable institutional influence over the Fund's core lending functions.

Second, contrary to much existing literature that emphasises American dominance, US influence shows a more selective pattern of impact. While American ties do not significantly affect loan volumes, participation rates, or approval decisions, they are strongly associated with *fewer conditions*, that is, a reduced number of performance criteria, benchmarks, and prior actions attached to lending arrangements. This finding suggests that US power manifests not through broad institutional control over access to resources, but through targeted influence over the design of conditionality. However, this result should not be interpreted as evidence of systematically “more lenient” conditionality: our additional analyses find no evidence that American ties reduce the required fiscal adjustment imposed by IMF programmes.

Third, our robustness checks reveal that European influence operates on a global rather than regional or colonial basis. Neither European borrowers nor African recipients benefit disproportionately from their ties to European powers. If anything, the evidence suggests these groups may receive less favourable treatment than other countries with comparable alignment profiles. This pattern indicates that European influence reflects genuine institutional power within the IMF rather than geographic or historical favouritism.

These findings have several important implications for our understanding of international financial institutions and global economic governance more broadly. Theoretically, they complicate simple narratives of American hegemony in international organisations. While the United States undoubtedly remains the most powerful single actor within the Bretton Woods system, our results suggest that European powers collectively exercise substantial influence over key institutional functions. This finding aligns with observations about the IMF's historical leadership structure—epitomised by the "gentlemen's agreement" reserving the Managing Director position for Europeans—and suggests that informal power arrangements may be more consequential than formal voting shares or veto rights.

The asymmetric nature of US and European influence also speaks to broader debates about the sources and exercise of institutional power. European influence appears to operate through channels that affect aggregate resource allocation—determining who receives assistance and in what amounts. American influence, by contrast, manifests through mechanisms that shape the policy content of lending arrangements—affecting what borrowers must do to receive and maintain access to funds. This division of labour may reflect different institutional strengths, strategic priorities, or coordination mechanisms between the two actors.

From a policy perspective, these findings have implications for ongoing debates about IMF reform and representation. Calls for enhanced emerging market voice and reduced Western dominance must grapple with the reality that "the West" is not a monolithic bloc. European and American interests may diverge in important ways, creating opportunities for coalition-building and institutional change that would not exist under unified Western control. Understanding these internal dynamics may be crucial for reform advocates seeking to reshape global financial governance.

Our analysis also sheds light on the nature of conditionality, one of the most controversial aspects of IMF operations. The finding that American influence is associated with fewer conditions, while European influence shows no such effect, suggests that critiques of overly intrusive conditionality may need to be more precisely targeted. European powers, despite their historical association with social democratic governance models, do not appear to advocate for more lenient policy requirements. This pattern may reflect the influence of German-style fiscal conservatism within European policy circles or the adoption of neoliberal economic orthodoxy by European technocratic elites.

Several limitations of our analysis should be acknowledged. First, our sample period ends in 2010, reflecting the coverage of the underlying datasets we rely on. While this prevents us from capturing the European sovereign debt crisis and the subsequent Troika arrangements, we expect that including the post-2010 period would, if anything, reinforce our findings. The IMF’s close collaboration with European institutions during the crisis suggests that European influence over lending decisions likely became even more pronounced after our period of observation.

Second, our measures of influence, while grounded in the existing literature, capture alignment rather than direct lobbying or pressure. We cannot observe the specific mechanisms through which influence is exercised or the content of behind-the-scenes negotiations. Qualitative case studies examining specific lending decisions could provide valuable complementary insights into the micro-foundations of the patterns we document. However, such approaches often rely on extensive insider information and interviews with IMF officials, which are difficult to obtain given the Fund’s highly consensus-based and closed-door decision-making culture.

Third, in terms of scope, three clarifications are in order. Our analysis focuses exclusively on lending outcomes and does not examine other dimensions of IMF operations, such as surveillance, technical assistance, or institutional governance. European and American influence may well manifest differently across these functions. Second, the effectiveness of IMF policies is also beyond the scope of our analysis. A large body of literature already addresses this question, and its findings can be combined with ours to inform policy debates. Third, our approach is deliberately empirical and positivist, aimed at identifying systematic patterns of influence rather than assessing the *appropriateness* of IMF lending arrangements. Questions of legitimacy and normative evaluation, long central to debates about the Fund, are not the direct focus of this paper, though we recognise that our results have implications for these debates and welcome further normative interpretations.

Looking ahead, several promising avenues for future research emerge from our findings. First, it is worth re-examining the determinants of IMF conditionality by moving beyond simple counts of conditions. In addition to the required fiscal adjustment, future work could measure conditionality across specific domains—such as trade liberalisation, privatisation, or structural reforms—to assess whether transatlantic influence shapes not only the quantity but also the substance of IMF policy requirements.

A second line of inquiry would be to examine the domestic sources of IMF influence. Our study measures international alignment, but the national actors transmitting influence remain a black box. Case studies focusing on finance ministries, central banks, foreign ministries, or private banking interests could shed light on which domestic constituencies mobilise influence within the IMF, and whether these channels differ between Europe and the United States.

Finally, our findings suggest the value of applying a similar disaggregation of Western influence to other international organisations. Beyond the World Bank, promising candidates include the OECD, the Bank for International Settlements (BIS), the WTO, and the European Bank for Reconstruction and Development (EBRD). Each involves overlapping Western leadership but potentially divergent US and European priorities, offering a fertile testing ground for whether the asymmetries we document are specific to the IMF or reflect broader patterns of transatlantic governance.

In conclusion, this paper demonstrates that the International Monetary Fund, despite its technocratic veneer, remains deeply shaped by geopolitical considerations. However, the nature of great power influence is more complex and differentiated than often assumed. European powers exercise broad institutional influence over access to IMF resources, while American power manifests more selectively through conditionality. This division of influence has important implications for understanding both the current operations of the international financial system and the prospects for its future reform. As emerging economies continue to challenge Western dominance in global governance, recognising the internal dynamics and potential fractures within the Western coalition becomes increasingly important for scholars, policymakers, and reform advocates alike.

@@

```{r}
#| label: fig-scree-eu
#| fig-cap: Scree Plot of European Influence PCA
#| fig-pos: H
load('save/scree.RData')
scree_EU
```

>>>>

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
date-format: |
    | DD MMMM YYYY
authors:
  - name: Dianyi Yang
    email: dianyi.yang@politics.ox.ac.uk
    orcid: 0009-0004-4652-3429
    corresponding: true
    affiliations:
      - ref: ox    
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
    affiliations:
      - ref: bw
    equal-contributor: true
  - name: Zhenyi Chen
    orcid: 0009-0008-2910-5187
    affiliations:
      - ref: rmu
    equal-contributor: true
affiliations:
  - id: ox
    name: Department of Politics and International Relations, University of Oxford
  - id: bw
    name: School of International Relations and Diplomacy, Beijing Foreign Studies University
  - id: rmu
    name: School of International Studies, Renmin University of China
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
          \AtAppendix{\counterwithin{table}{section}}
abstract: "This paper investigates whether the United States or Europe holds greater influence over the International Monetary Fund (IMF)—a longstanding yet underexamined question in the study of global economic governance. While the Fund is widely regarded as dominated by “Western” powers, most existing research either focuses exclusively on American influence or treats the West as a unified bloc, masking important variation between its two key actors. Drawing on panel data from 1980 to 2010, we construct separate measures of US and European influence using principal component analysis of commonly used alignment indicators—UN voting patterns, bilateral trade, and banking exposure—and estimate a series of Tobit and Probit models to assess their impact on four key IMF lending outcomes: loan-to-GDP ratios, participation rates, loan approvals, and conditionality. We find that European influence is strongly associated with higher loan volumes, more frequent participation, and greater likelihood of loan approval, while US influence is not. However, American ties are significantly correlated with fewer conditions attached to loans. Additional analysis shows that European influence is not driven by regional or colonial favouritism: neither European nor African borrowers benefit more from European ties—and may, in fact, fare worse—than other countries with comparable alignment. These findings complicate existing assumptions about IMF decision-making, highlighting the need to distinguish between types of Western influence. They also suggest that Europe wields broader institutional influence over access to IMF resources, while US power manifests more selectively through conditionality. In doing so, the paper contributes to debates on hegemony, institutional legitimacy, and intra-Western dynamics in global economic governance."
keywords: IMF, US hegemony, Europe
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, South Korea, and MENA countries [@vreeland2009; @copelovitch2010a; @lipscy2018; @harrigan2006]. Closer allies of the United States enjoy IMF loans with less fiscal conditions [@dreher2003]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

@@

In principle, quantitative approaches are well suited for directly comparing the influence of different great powers within international organisations. In the IMF context, however, most such studies focus on American influence, with little systematic attention paid to European power. These studies typically proxy the strategic interests of major powers using indicators such as UN voting alignment [@thacker1999; @dreher2009], bilateral trade intensity [@barro2005], and financial exposure [@oatley2004]. Yet, much of this literature has remained US-centric, often using only American alignment as the independent variable of interest [@thacker1999; @oatley2004; @dreher2009; @guimaraes2021]. Other studies pool European indicators with US ones under the broader category of “Western” influence, thus masking potential differences between the two [@lipscy2018; @barro2005; @youssef2023; @guimaraes2021][^2]. These ties have been shown to correlate with more favourable IMF lending outcomes—larger loans, higher approval rates, and fewer or less stringent conditions [@lipscy2018; @barro2005; @dreher2009; @youssef2023; @guimaraes2021]. Yet, despite the availability of these tools, comparative assessments between the US and Europe remain largely absent from this literature.

[^2]: @youssef2023 also include China as a major shareholder when calculating political influence.

@@

-   UN General Assembly voting similarity[^5]
-   Bilateral trade intensity[^6]
-   Banking sector exposure[^7]

[^5]: Originally used in @thacker1999.

[^6]: From @barro2005.

[^7]: See @broz2006; @lipscy2017; @oatley2004.

```{r percentage_variance}
load('save/PCA.Rdata')
US_PC1_percentage <- US$sdev[1]^2 / sum(US$sdev^2) * 100
EU_PC1_percentage <- EU$sdev[1]^2 / sum(EU$sdev^2) * 100
```

However, as noted by @lipscy2018 and @belsley2004, including multiple proxies for the same latent construct in a single regression model leads to multicollinearity, which reduces statistical power and inflates standard errors, as the variables are highly correlated in our data[^8]. To address this issue, we apply principal component analysis (PCA) to reduce the dimensionality of the data [@dunteman2008]. Specifically, we extract the first principal component — a linear combination of the original variables that captures the greatest share of variance — to serve as a single measure of influence[^9].

[^8]: See Appendix [-@sec-corr].

[^9]: The first principal components capture the majority of the variance (`r format(EU_PC1_percentage, digits = 1)`% for Europe and `r format(US_PC1_percentage, digits = 1)`% for the United States). See Appendix [-@sec-scree] for scree plots and additional details.

@@

We include a range of economic performance indicators that have been shown to systematically affect IMF lending decisions [@barro2005; @knight1997; @youssef2023]. To capture these dynamics, we include GDP, per capita GDP, and lagged GDP growth as proxies for macroeconomic performance. International reserves as a share of imports are used to measure a country’s external liquidity and ability to manage balance-of-payments pressures without IMF support. Finally, a binary variable for OECD membership serves as a proxy for broader institutional and economic differences among countries.

@@

[^10]: Quadratic terms are included for GDP and per capita GDP to capture potential non-linear effects. See [Model Specification](#sec-model) for details.
## Summary Statistics

@@

# Case Study

@@

@@

```{r}
#| label: fig-scree-eu
#| fig-cap: Scree Plot of European Influence PCA
#| fig-pos: H
load('save/scree.Rdata')
scree_EU
```
